Understanding the Impact of Mortgage State and Income on Household Poverty Status

### 2024-04-19 paper/US_Poverty_Analysis_2019.qmd
final edits

The logistic regression model shows that in the US in 2019, households with an annual total income less than $50000 are almost all in poverty. For households earning $50000 to $100000, the majority is not in poverty, yet a notable amount of them are still in poverty. For households earning above $100000, all of them are not in poverty. Additionally, renters are more likely to be in poverty compared to owners with or without mortgage.

@@

## Relationship betwen Mortgage State and Poverty Status
In the USA in 2019, housing costs accounted for more than 30% of household expenditure, making mortgage status a critical factor in determining a family's risk of falling into poverty [@Statistics_2020]. The percentage of individuals below the poverty threshold, adjusted for housing costs are: $29,080 for a family of four with a mortgage, $24,413 for a family of four without a mortgage, and $29,194 for a family of four renting [@Khattar_Pathak_Schweitzer_Khan_McManus_Jones_Benson_Kelly_2023]. From this, we can see that families without mortgages experienced the lowest housing costs, leaving more funds available for other essentials. In contrast, families with mortgages may face expensive monthly payments, especially if they have chosen a floating interest rate, making them vulnerable to rising interest rates. Renters also experience challenges when the rent increases. So they either have to accept higher housing costs or waste time and energy to find new homes. This supports my model result that renters are more likely to be in poverty.

Also, low-income households have a harder time to access lower mortgage rates and reduce debt burdens through refinancing. Households with low to moderate incomes were less able to take advantage of lower mortgage rates and relieve their debt burdens through refinancing during the period of exceptionally low rates: only 24% of mortgages in low-income regions underwent refinancing, in contrast to the 42% in high-income areas [@mortgage]. Therefore, it is generally harder for low-income households to repay their mortgage than higher income households.

These problems with housing make it more likely for families which are already in poverty, to become poorer. So it is important to make housing more affordable for people, in order to help people out of poverty.

@@

## US Government Programs to Help
For household in poverty, escaping poverty is tough, but the US government and communities are stepping up to help. For example, single parents often struggle with low incomes. But they can get help from government programs such as the Supplemental Nutrition Assistance Program (SNAP) [@snap], commonly known as food stamps, which provides eligible low-income individuals and families with funds to purchase nutritious food. They can also receive child support payments to help cover expenses for their kids. Families with low incomes and lots of kids can also get help. Programs like the Child Tax Credit (CTC) [@ctc], which provides tax credits to eligible families with qualifying children. The credit currently provides up to $2,000 per child to about 40 million families every year []. So obviously, these programs are a huge help for low-income family with kids. Other programs, like housing assistance and Medicaid, provide support with things like affordable housing and healthcare. These programs are all aimed at helping families in need to get back on their feet and improve their poverty situation.

@@

### Lump sum nature of tax credit
In the SPM, taxes paid and tax credits in the year that are earned are include [@spm]d. But typically the credit would not appear in the family income until the next year. Also, tax credits are given as a lump sum, but poverty is a something that might happen anytime in the year. For example, a family may experience poverty because someone lost a job and cannot find another, so the family is in poverty. The tax credit may or may not provide relief depending on the timing in the year and the severity of expenses faced by the family throughout the year. So it is not very accurate that the SPM examine resources in a full calendar year compared with a threshold based on the calendar year. The family might have experienced 11 months of poverty before they got their tax credits in the 12th month.

>>>>

The logistic regression model shows that in the US in 2019, households with an annual total income less than $5000 are in poverty. For households earning $5000 to $50000, the majority is not in poverty, yet some of them are still in poverty. For households earning above $100000, all of them are not in poverty. Additionally, renters are more likely to be in poverty compared to owners with or without mortgage.

@@

## Relationship betwen Mortgage State and Poverty Status
In the USA in 2019, housing costs accounted for more than 30% of household expenditure, making mortgage status a critical factor in determining a family's risk of falling into poverty [@Statistics_2020]. A study shows that in 2019, the annual income needed are: $29,080 for a family of four with a mortgage; $24,413 for a family of four without a mortgage; and $29,194 for a family of four renting [@Khattar_Pathak_Schweitzer_Khan_McManus_Jones_Benson_Kelly_2023]. From this, we can see that families without mortgages experienced the lowest housing costs, leaving more funds available for other essentials. In contrast, families with mortgages may face expensive monthly payments, especially if they have chosen a floating interest rate, making them vulnerable to rising interest rates. Renters also experience challenges when the rent increases. So they either have to accept higher housing costs or waste time and energy to find new homes. This supports my model result that renters are more likely to be in poverty, followed by owners with mortgage, then owners without mortgage.

Also, low-income households have a harder time to access lower mortgage rates and reduce debt burdens through refinancing. Households with low to moderate incomes were less able to take advantage of lower mortgage rates and relieve their debt burdens through refinancing during the period of exceptionally low rates: only 24% of mortgages in low-income regions underwent refinancing, in contrast to the 42% in high-income areas [@mortgage]. Therefore, it is generally harder for low-income households to repay their mortgage than higher income households. The challenge of paying mortgage might increase household's likelihood to be in poverty. We can see that these problems with housing make it more likely for families which are already in poverty, to become poorer. So it is important to make housing more affordable for people, in order to help people out of poverty.

@@

## US Government Programs to Help
For household in poverty, escaping poverty is tough, but the US government and communities are stepping up to help. For example, single parents often struggle with low incomes. But they can get help from government programs such as the Supplemental Nutrition Assistance Program (SNAP) [@snap], commonly known as food stamps, which provides eligible low-income individuals and families with funds to purchase nutritious food. They can also receive child support payments to help cover expenses for their kids. Families with low incomes and lots of kids can also get help. Programs like the Child Tax Credit (CTC) [@ctc], which provides tax credits to eligible families with qualifying children. The credit currently provides up to $2,000 per child to about 40 million families every year [@child]. So obviously, these programs are a huge help for low-income family with kids. Other programs, like housing assistance and Medicaid, provide support with things like affordable housing and healthcare. These programs are all aimed at helping families in need to get back on their feet and improve their poverty situation.

@@

### Lump sum nature of tax credit
In the SPM, taxes paid and tax credits in the year that are earned are included [@spm]. But typically the credit would not appear in the family income until the next year. Also, tax credits are given as a lump sum, but poverty is a something that might happen anytime in the year. For example, a family may experience poverty because someone lost a job and cannot find another, so the family is in poverty. The tax credit may or may not provide relief depending on the timing in the year and the severity of expenses faced by the family throughout the year. So it is not very accurate that the SPM examine resources in a full calendar year compared with a threshold based on the calendar year. The family might have experienced 11 months of poverty before they got their tax credits in the 12th month.
